---
title: "Project 2: A minimal I-map algorithm"
author: "王纪元 Wang Jiyuan"
date: "2024/05/23"
---

## Exercise 1

```{r}
# distGlobal <- read.delim(file.choose(), header = TRUE)
distGlobal <- read.delim("global_distribution.txt", header = TRUE)

print(distGlobal)
```

```{r}
print(class(distGlobal))

print(class(distGlobal$D))

print(class(distGlobal$I))

print(class(distGlobal$G))

print(class(distGlobal$S))

print(class(distGlobal$L))

print(class(distGlobal$probability))
```

## Exercise 2

```{r}
# Input parameters:
#   2 character type vectors, A and B, which will represent both sets.

# Output Parameters:
#   A boolean value, isSubset, which will be TRUE if A is a strict subset of
#   B (A ⊂ B); or FALSE if A is not strict subset of B.
IsSubset <- function(A, B) {
  isSubset <- all(A %in% B) && length(A) < length(B)
  return(isSubset)
}

# Test the function
A <- c("a", "b", "c")
B <- c("a", "b", "c", "d")

print(IsSubset(A, B))

A <- c("A", "b", "c")
B <- c("a", "b", "c", "d")

print(IsSubset(A, B))

A <- c("a", "b", "c", "d")
B <- c("a", "b", "c", "d")

print(IsSubset(A, B))

A <- character(0)
B <- c("a", "b", "c", "d")

print(IsSubset(A, B))
```

## Exercise 3

```{r}
# Load the required libraries
library(rlist)
library(dplyr)
```

`IsCondIndependent`, which will constitute the main core of the minimal Imap 
algorithm. This function should be able to verify 2 types of conditional
independencies within a global distribution P.

Input Parameters:

- **P**: Data frame of the global distribution P (with the format of
distGlobal: probabilities in the last column).
- **X**: Name of the variable X within the distribution P.
- **Y**: Character type vector Y, with variable names Y1, . . ., Ym within distribution P.
- **Z**: Vector of character type Z, with variable names Z1, . . ., Zl

within distribution P. If the vector Z is the empty set (character(0)),
IsCondIndependent will only check if (X ⊥ Y). If the vector Z is
not the empty set, IsCondIndependent will only check if (X ⊥ Y | Z).

Output Parameters:

A boolean value, independent, which will be
TRUE if in the distribution P it is satisfied that (X ⊥ Y | Z) (or X ⊥ Y,
if Z = character(0)); or it will be FALSE if in the distribution P it is
not satisfied that (X ⊥ Y | Z) (or X ⊥ Y, if Z = character(0)).

```{r}
IsCondIndependent <- function(P, X, Y, Z) {
  # Initialize
  independent <- TRUE

  get_levels <- function(X) {
    X_index <- match(X, colnames(P))
    X_factor <- as.factor(P[, X_index])
    X_levels <- levels(X_factor)
    return(X_levels)
  }


  levels_ <- list()
  levels_ <- list.append(levels_, get_levels(X))

  for (i in seq_along(Y)) {
    levels_ <- list.append(levels_, get_levels(Y[i]))
  }

  for (i in seq_along(Z)) {
    levels_ <- list.append(levels_, get_levels(Z[i]))
  }

  # print(levels_)

  combinations <- expand.grid(levels_, stringsAsFactors = FALSE)

  # print(combinations)

  for (i in seq_len(dim(combinations)[1])) {
    if (length(Z) == 0) {
      # X ⊥ Y
      vectorXY <- combinations[i, ]
      vectorX <- vectorXY[1]
      vectorY <- vectorXY[2:length(vectorXY)]

      # print("vectorXY:")
      # print(vectorXY)
      # print("vectorX:")
      # print(vectorX)
      # print("vectorY:")
      # print(vectorY)

      probXY <- 0
      probX <- 0
      probY <- 0

      for (j in seq_len(dim(P)[1])) {
        rowP <- P[j, ]
        prob <- rowP[length(rowP)]
        # print("rowP:")
        # print(rowP)
        # print("prob:")
        # print(prob)
        if (IsSubset(vectorXY, rowP)) {
          probXY <- probXY + prob
        }
        if (IsSubset(vectorX, rowP)) {
          probX <- probX + prob
        }
        if (IsSubset(vectorY, rowP)) {
          probY <- probY + prob
        }
      }
      if (!near(probXY, probX * probY)) {
        independent <- FALSE
        break
      }
    } else {
      # X ⊥ Y | Z
      vectorXYZ <- combinations[i, ]
      vectorX <- vectorXYZ[1]
      vectorY <- vectorXYZ[2:(length(vectorXYZ) - length(Z))]
      vectorZ <- vectorXYZ[(length(vectorXYZ) - length(Z) + 1):length(vectorXYZ)]

      probXYZ <- 0
      probXZ <- 0
      probYZ <- 0
      probZ <- 0

      for (j in seq_len(dim(P)[1])) {
        rowP <- P[j, ]
        prob <- rowP[length(rowP)]
        if (IsSubset(vectorXYZ, rowP)) {
          probXYZ <- probXYZ + prob
        }
        if (IsSubset(vectorX, rowP) && IsSubset(vectorZ, rowP)) {
          probXZ <- probXZ + prob
        }
        if (IsSubset(vectorY, rowP) && IsSubset(vectorZ, rowP)) {
          probYZ <- probYZ + prob
        }
        if (IsSubset(vectorZ, rowP)) {
          probZ <- probZ + prob
        }
      }
      if (!near(probXYZ * probZ, probXZ * probYZ)) {
        independent <- FALSE
        break
      }
    }
  }
  return(independent)
}
```

The function returns `TRUE` in the following conditional independencies:

- (D ⊥ I)
- (D ⊥ S)
- (G ⊥ S | I, D)
- (S ⊥ D | I)
- (I ⊥ L | G, S)
- (D ⊥ L, S | I, G)

The function returns `FALSE` in the following conditional independencies:

- (G ⊥ L)
- (G ⊥ S)
- (S ⊥ L)
- (S ⊥ L, D, G)
- (I ⊥ S | G, L)
- (I ⊥ G | S, L)

```{r}
# Test the function
if (IsCondIndependent(distGlobal, "D", c("I"), character(0)) != TRUE) {
  print("Test failed for (D ⊥ I)")
}

if (IsCondIndependent(distGlobal, "D", c("S"), character(0)) != TRUE) {
  print("Test failed for (D ⊥ S)")
}

if (IsCondIndependent(distGlobal, "G", c("S"), c("I", "D")) != TRUE) {
  print("Test failed for (G ⊥ S | I, D)")
}

if (IsCondIndependent(distGlobal, "S", c("D"), c("I")) != TRUE) {
  print("Test failed for (S ⊥ D | I)")
}

if (IsCondIndependent(distGlobal, "I", c("L"), c("G", "S")) != TRUE) {
  print("Test failed for (I ⊥ L | G, S)")
}

if (IsCondIndependent(distGlobal, "D", c("L", "S"), c("I", "G")) != TRUE) {
  print("Test failed for (D ⊥ L, S | I, G)")
}

if (IsCondIndependent(distGlobal, "G", c("L"), character(0)) != FALSE) {
  print("Test failed for (G ⊥ L)")
}

if (IsCondIndependent(distGlobal, "G", c("S"), character(0)) != FALSE) {
  print("Test failed for (G ⊥ S)")
}

if (IsCondIndependent(distGlobal, "S", c("L"), character(0)) != FALSE) {
  print("Test failed for (S ⊥ L)")
}

if (IsCondIndependent(distGlobal, "S", c("L", "D", "G"), character(0)) != FALSE) {
  print("Test failed for (S ⊥ L, D, G)")
}

if (IsCondIndependent(distGlobal, "I", c("S"), c("G", "L")) != FALSE) {
  print("Test failed for (I ⊥ S | G, L)")
}

if (IsCondIndependent(distGlobal, "I", c("G"), c("S", "L")) != FALSE) {
  print("Test failed for (I ⊥ G | S, L)")
}
```

## Exercise 4

```{r}
library(DescTools)
library(Rgraphviz)
library(bnlearn)
```

```{r}
# This function will make use of the functions already developed: IsSubset and 
# IsCondIndependent.

# Input Parameters:
# • P:Data frame of the global distribution P (with the format of
# distGlobal: probabilities in the last column).
# • Xord: Vector of type character that will contain all the variables
# of P in a certain order: X1, X2, . . . , Xn.

# Output Parameters:
# A minimal I-map G (associated to P and Xord). G shall be a list consisting of
# 2 elements:
# 1. A character type vector, variables, which will simply be equal
# to Xord.
# 2. A character type array, arcs, which will hold the relationships
# between the Xord variables detected by the algorithm. Each row
# of arcs will have a relationship of the form A → B. Therefore,
# arcs will have 2 columns: first column for the parent variable of
# the relationship (A) and second column for the child variable of
# the relationship (B).
MinimalIMap <- function(P, Xord) {
  n <- length(Xord)
  vectorArcs <- c()

  for (i in 2:n) {
    U <- Xord[1:(i - 1)]
    for (j in (i - 1):0) {
      matrixUfirst <- CombSet(U, j)
      for (k in seq_len(dim(matrixUfirst)[1])) {
        Ufirst <- matrixUfirst[k, ]
        Dif <- setdiff(Xord[1:(i - 1)], Ufirst)
        if (IsSubset(Ufirst, U)) {
          if (IsCondIndependent(P, Xord[i], Dif, Ufirst)) {
            U <- Ufirst
          }
        }
      }
    }
    if (length(U) >= 1) {
      # print(U)
      for (j in seq_along(U)) {
        relationship <- c(U[j], Xord[i])
        vectorArcs <- c(vectorArcs, relationship)
      }
    }
  }
  arcs <- matrix(vectorArcs, ncol = 2, byrow = TRUE)
  G <- list(variables = Xord, arcs = arcs)
  return(G)
}
```

```{r}

# If all went well, run MinimalIMap with P = distGlobal, testing 3 variable
# orders: Xord = c("D", "I", "S", "G", "L"), Xord = c("L", "S", "G",
# "I", "D") and Xord = c("L", "D", "S", "I", "G").

testcase <- function(Xord) {
  G <- MinimalIMap(distGlobal, Xord)
  dag <- empty.graph(G$variables)
  arcs(dag) <- G$arcs
  graphviz.plot(dag)
}

testcase(c("D", "I", "S", "G", "L"))

testcase(c("L", "S", "G", "I", "D"))

testcase(c("L", "D", "S", "I", "G"))
```

## Conclusions

This project took me approximately 4 hours to complete. I learned how to
implement function `IsSubset` to verify if a set is a strict subset of another
set, and function `IsCondIndependent` to verify conditional independencies
within a global distribution P. I also learned how to implement a minimal I-map 
algorithm using the functions `IsSubset` and `IsCondIndependent`. I tested the
algorithm with the global distribution `distGlobal` and 3 variable orders. The
algorithm successfully detected the relationships between the variables in the
global distribution.


